---
title: "Day 1: Historian Hysteria"
author: "Celine Wuyts"
date: today
date-format: long
format: 
  pdf: 
    documentclass: scrreport
    execute: 
      echo: false
      warning: false
      error: false
  gfm:
    execute: 
      echo: true
      warning: false
      error: false
  html:
    theme: 
      - flatly
      - styles.scss
    toc: true
    self-contained: TRUE
    code-fold: true
    code-tools: true
    execute: 
      echo: true
      warning: false
      error: false
---



```{r load_libraries}

library(tidyverse)
```

```{r get_labels, echo = FALSE}

labs <- knitr::all_labels()
labs_0 <- str_subset(labs, "^part_0-")
labs_1 <- str_subset(labs, "^part_1-")
labs_2 <- str_subset(labs, "^part_2-")

labs_1 <- c(labs_0, labs_1)
labs_2 <- c(labs_0, labs_2)
```

{{< pagebreak >}}

::: {.callout-tip}
## Attribution

This is an R solution for [Advent of Code 2024, Day 1](https://adventofcode.com/2024/day/1).

:::

# Part 1

The puzzle input has two columns with location ID numbers.

```{r part_0-read_data}

dat <- read.table("data/day-1_input.txt", header = FALSE, sep = "")
```

```{r}

glimpse(dat)
```

Let's sort each column and add a difference variable.

```{r part_1-compute_rowwise_distance}

dat_sorted <- map_df(dat, ~ sort(.))
dat_sorted <- dat_sorted %>% mutate(diff = abs(V1 - V2))
```
```{r}

glimpse(dat_sorted)
```

Let's compute the total distance between the left list and the right list.

```{r part_1-compute_total_distance}

total_diff <- dat_sorted %>% pull(diff) %>% sum()
```

The total distance is `r total_diff`.

::: {.content-visible when-format="pdf"}

## R code

```{r ref.label=labs_1}
#| eval: false
#| echo: true
```

::: 

# Part 2

Instead, let's count how often each number from the left list appears in the right list, and add a "similarity" variable by multiplying each number in the left list by the number of times that number appears in the right list.

```{r part_2-compute_rowwise_similarity}

dat_counts <- dat %>% count(V1, name = "n_V1")

dat_counts <- dat_counts %>% 
  mutate(n_V2 = map_int(V1, ~ sum(.x == dat$V2)),
         similarity = V1 * n_V2)
```

```{r}

glimpse(dat_counts)
```

Let's compute the total similarity score between the left list and the right list.

```{r part_2-compute_total_similarity}

total_similarity <- dat_counts %>% pull(similarity) %>% sum()
```

The total similarity score is `r total_similarity`.

::: {.content-visible when-format="pdf"}

## R code

```{r ref.label=labs_2}
#| eval: false
#| echo: true
```

::: 

# Session information

```{r close_out}

sessionInfo()
```

